HO unification from object language to meta language

### 2024-04-03 paper/main.tex
todo in diff colors

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.
The code discussed in the paper can be accessed at the URL:
\url{https://github.com/FissoreD/paper-ho}.

@@

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
TODO extend \Eo and \Ee with reflexivity on uvars.

@@

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress XXX improve....

@@

\todo[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\todo[inline]{is new used?}

@@

TODO: add ref to \cref{sec:invariant1}

@@

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\todo[inline]{explain better}

@@

% The solution we are proposing aim to overcome these unification issues by 1)
% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
% the variables of the OL with the decompiled version of their corresponding meta
% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
% and that the substitution in the object language is the same as the one returned
% by the ML. \todo[inline]{same or $\supseteq$ or $\subseteq$}

@@

\subsection{Compilation}
\label{sec:compilation}
\todo[inline]{manca beta normal in entrata}

@@

% The following code
% %
% \begin{elpicode}
%   kind link type.
%   type link nat -> nat -> nat -> subst.
% \end{elpicode}
% %
% \noindent
% defines a link, which is a relation between to variables indexes, the first
% being the index of a OL variable and the second being the index of a ML
% variable. The third integer\todo[inline]{integer or nat?} is the number of term in the
% scope of the two variables, or equivalently, in a typed language, their arity.

@@

\todo[inline]{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

@@

\noindent
Note that this condition does not require the term to be in \llambda.
\todo[inline]{Is this relevant}

@@

\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. There are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$  (under
    the context $\Gamma$)
  \item if $X = \lambda x.t$ we run $X \Ue{} T$.
  \todo[inline]{do we have the name case in the code?}
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\todo[inline]{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

\paragraph{Example of \progressetaright}\todo[inline]{TODO}

A second example, showing the activation of a link when the \rhs is no more a
\maybeeta, is given in \cref{sec:invariant1}, since we need to work with
variables used with different arities. This example represent the run of the
unification problems proposed at \cref{sec:basic-comp-limitations}

@@

\todo[inline]{I don't get this TODO}
TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
not reduce to $T x$ where x is not free in $T$.

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\todo[inline]{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

\todo[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
}
\section{Handling of \maybebeta}\label{sec:beta}

@@

\todo[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. In general, unification in \maybebeta admits more then one
solution and commiting one of them in the substitution does not guarantee
prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\todo[inline]{it is even a ground term, there is no unification left to perform actually}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifable problem $(\lambda x. B) \appsep a \Ue a$.

@@

\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
% ... where S = [s1 .. sn] e L = [l1 .. lm]}
\todo[inline]{Is it clearer?}

@@

% \linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
% \todo[inline]{un po' ridondante con 8.2, non so se serva}

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the original unification problem is not in \llambda
  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  %\todo[inline]{Is it clearer?}
  and the unification fails.
  \todo[inline]{it is clear. Quello che manca è percè serve farlo, e che cosa si guadagna. Qui dimostri che fallire "ha senso" se siamo solo interessati a llambda, ma allo stesso tempo sti beta link sono lì per gestire roba fuori.}
\end{corollary}

@@

\begin{proof}[Proof sketch]
  By contradiction,\todo{we assume to have a link} if $m = 0$, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.\todo{since we started by assuming we had a link}
\end{proof}

@@

\todo[inline]{Dire che maybe eta fa il detect anche su 
termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
esempio?}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  %\todo[inline]{Is it clearer?}
  \todo[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
\end{definition}

@@

\begin{proof}[Proof sketch]
  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  creates each time a new \linketa and replace the old \linkbeta with a new one.
  This progression can however triggered at most $m$ times, since each new
  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  the \linkbeta is removed from \linkStore which is now filled by $m$ new
  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  terminates if \linkStore is made by only \linketa and therefore
  \elpiIn{progress} still terminates.
  %\todo[inline]{is it ok?}
  \todo[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
  \todo[inline]{is it clearer?}
\end{corollary}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers progress \todo{which rule?} in first \linketa that is 
removed and $E$ is assigned to $\lambda x.x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the following two links:
\linkbetaM{x}{C_x}{F_x} and
\todo{ma allora il link ora sparisce perhè è in llambda?}
\linketaM{}{E}{\lambda x.D_x}. The second unification
problem assignes $a$ to $B$, that in turn activates
the second \linketa ($a$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto a\}$.\todo{assegnament per D incompleto}
\todo{alla fine non si usa anche la regola beta-rigid-head?}

@@

\subsection{Tricky examples}

\begin{elpicode}
  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
  triple ok @X (@lam x\ x),
  triple ok @Y @f
\end{elpicode}

@@

\section{First order approximation}

\otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
\otext{We can re-use part of the algo for $\beta$ given before}

\section{Unif encoding in real life}
\otext{Il ML presentato qui è esattamente elpi}
\otext{Il OL presentato qui è esattamente coq}
\otext{Come implementatiamo tutto ciò nel solver}

\section{Results: stdpp and tlc}
\otext{How may rule are we solving?}
\otext{Can we do some perf test}

>>>>

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. \noteE[inline]{citare Teyjus}.
The code discussed in the paper can be accessed at the URL:
\url{https://github.com/FissoreD/paper-ho}.

@@

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}

@@

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress \noteE[inline]{XXX improve....}

@@

\noteE[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\noteE[inline]{is new used?}

@@

\noteD{add ref to \cref{sec:invariant1}}

@@

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}

@@

% The solution we are proposing aim to overcome these unification issues by 1)
% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
% the variables of the OL with the decompiled version of their corresponding meta
% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
% and that the substitution in the object language is the same as the one returned
% by the ML. \noteE[inline]{same or $\supseteq$ or $\subseteq$}

@@

\subsection{Compilation}
\label{sec:compilation}
\noteE[inline]{manca beta normal in entrata}

@@

% The following code
% %
% \begin{elpicode}
%   kind link type.
%   type link nat -> nat -> nat -> subst.
% \end{elpicode}
% %
% \noindent
% defines a link, which is a relation between to variables indexes, the first
% being the index of a OL variable and the second being the index of a ML
% variable. The third integer\noteE[inline]{integer or nat?} is the number of term in the
% scope of the two variables, or equivalently, in a typed language, their arity.

@@

\noteE[inline]{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

@@

\noindent
Note that this condition does not require the term to be in \llambda.
\noteE[inline]{Is this relevant}

@@

\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. There are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$  (under
    the context $\Gamma$)
    \noteE[inline]{where y comes from? $X = y$: $y$ is in the ctx of X}
    \item if $X = \lambda x.t$ we run $X \Ue{} T$.
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\noteE[inline]{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

@@

We delay in next section a \linketa progression due to \progressetaright

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

\noteD[inline]{ci vuole l'esempio per il \progressetaright}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

\noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
}
\section{Handling of \maybebeta}\label{sec:beta}

@@

\noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. In general, unification in \maybebeta admits more then one
solution and commiting one of them in the substitution does not guarantee
prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifable problem $(\lambda x. B) \appsep a \Ue a$.

@@

\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
% \noteE[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
% ... where S = [s1 .. sn] e L = [l1 .. lm]}
\noteE[inline]{Is it clearer?}

@@

% \linkbeta are put in \linkStore and activated when \rhs falls in \llambda. 
% \noteE[inline]{un po' ridondante con 8.2, non so se serva}

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the original unification problem is not in \llambda
  % \noteE[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  %\noteE[inline]{Is it clearer?}
  and the unification fails.
  \noteE[inline]{it is clear. Quello che manca è percè serve farlo, e che cosa si guadagna. Qui dimostri che fallire "ha senso" se siamo solo interessati a llambda, ma allo stesso tempo sti beta link sono lì per gestire roba fuori.}
\end{corollary}

@@

\begin{proof}[Proof sketch]
  By contradiction,\noteE{we assume to have a link} if $m = 0$, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.\noteE{since we started by assuming we had a link}
\end{proof}

@@

\noteE[inline]{Dire che maybe eta fa il detect anche su 
termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
esempio?}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  %\noteE[inline]{Is it clearer?}
  \noteE[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
\end{definition}

@@

\begin{proof}[Proof sketch]
  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  creates each time a new \linketa and replace the old \linkbeta with a new one.
  This progression can however triggered at most $m$ times, since each new
  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  the \linkbeta is removed from \linkStore which is now filled by $m$ new
  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  terminates if \linkStore is made by only \linketa and therefore
  \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the variables occurring in its \rhs are in \llambda.
  % \noteE[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
  \noteE[inline]{is it clearer?}
\end{corollary}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers progress \noteE{which rule?} in first \linketa that is 
removed and $E$ is assigned to $\lambda x.x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the following two links:
\linkbetaM{x}{C_x}{F_x} and
\noteE{ma allora il link ora sparisce perhè è in llambda?}
\linketaM{}{E}{\lambda x.D_x}. The second unification
problem assignes $a$ to $B$, that in turn activates
the second \linketa ($a$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A = \lambda x.x, B = a,
C_x = a, D = \lambda \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto a\}$.\noteE{assegnament per D incompleto}
\noteE{alla fine non si usa anche la regola beta-rigid-head?}

@@

\subsection{Heuristics}

Possiamo perdere OC e rimanere sospesi o commit di soluzione approx
\begin{elpicode}
  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
  triple ok @X (@lam x\ x),
  triple ok @Y @f
\end{elpicode}

@@

% \section{First order approximation}

% \otext{Coq can solve this: \coqIn{f 1 2 = X 2}, by setting X to f 1}
% \otext{We can re-use part of the algo for $\beta$ given before}

% \section{Unif encoding in real life}
% \otext{Il ML presentato qui è esattamente elpi}
% \otext{Il OL presentato qui è esattamente coq}
% \otext{Come implementatiamo tutto ciò nel solver}

% \section{Results: stdpp and tlc}
% \otext{How may rule are we solving?}
% \otext{Can we do some perf test}

\section{Practical example}
